OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-04 paper/paper.Rmd
Bold places where I need to insert exact numbers

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. Here, we tested the OptiFit algorithm with the reference as databases or _de
novo_ OTUs and compared the performance to existing tools.

@@

@@

- **TODO:** separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median MCC values.
- **TODO:** avoid slightly/much -- use numbers, e.g. X% better/worse.
- **TODO:** put the comparison at the beginning of the sentence, so you don't have to know what it is at the end.

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). In open reference mode, OTU quality was similar
between fitting the datasets to reference OTUs with OptiFit and clustering the
datasets _de novo_ with OptiClust. This held true for all datasets and reference
databases. However, in closed reference mode, OTU quality was slightly worse
when fitting to Greengenes and Silva, and much worse when fitting to RDP as
compared to OptiClust. No more than half of query sequences were fit to
reference OTUs in closed reference mode across any dataset/database combination.
**TODO: specify exact numbers for fraction mapped.**
OptiFit was able to fit **X** more query sequences to reference OTUs created with the
Greengenes and Silva databases than with RDP. In terms of run time,
closed reference OptiFit outperformed OptiClust, while OptiClust outperformed
open reference OptiFit.

To compare to existing software, vsearch was used to cluster OTUs _de novo_ or
with reference-based clustering to the greengenes database. For all datasets and
clustering methods (_de novo_, open reference, and closed reference), mothur's
clustering algorithms produced higher quality OTUs than vsearch. When
closed reference clustering against the greengenes database, vsearch was able to
map more query sequences to the reference than mothur's OptiFit algorithm. In
terms of runtime, OptiFit generally performed faster than vsearch when reference
clustering, while vsearch _de novo_ clustering outperformed OptiClust.

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode. OTU quality was remarkably stable across 100 different random seeds. In
terms of runtime, closed reference OptiFit performed faster than OptiClust on
whole datasets. In open reference mode, OptiClust performed faster than OptiFit
only when the OptiFit reference fraction was 30% or less. The split dataset
strategy performed just as well as the database strategy in open reference mode
regardless of database used, and outperformed the database strategy in
closed reference mode.

We also tested three methods for selecting the sequences to be used as the
reference; a simple random sample, weighting sequences by relative abundance,
and weighting by similarity to other sequences in the dataset. OTU quality was
similar with the simple and abundance-weighted sampling, but slightly worse with
similarity-weighted sampling. In closed reference mode, The fraction of query
sequences that can be fit to the reference OTUs decreases as the reference
fraction increases.

>>>>

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. **TODO: more words here?** 
Here, we tested the OptiFit algorithm with the reference as databases or _de
novo_ OTUs and compared the performance to existing tools.

@@

**TODO: make a plan for the figures**

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). As a comparison to existing software, vsearch was
used to cluster OTUs _de novo_ or with reference-based clustering to the
greengenes database. In open reference mode, OTU quality was similar between
fitting the datasets to reference OTUs with OptiFit and clustering the datasets
_de novo_ with OptiClust (**TODO: diff. in medians**). This held true for all
datasets and reference databases. OptiFit produced higher quality OTUs than
vsearch when open reference clustering against the greengenes database, with
median MCC scores of **A** and **B%** (respectively).
**TODO: open ref runtime of vsearch vs optifit vs opticlust**.

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and Silva, and **Y%** worse when
fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
in closed reference mode across any dataset/database combination. OptiFit was
able to fit **X** more query sequences to reference OTUs created with the
Greengenes and Silva databases than with RDP. vsearch was able to map more query
sequences to the greengenes reference than OptiFit in closed reference mode. In
terms of run time, closed reference OptiFit outperformed OptiClust by **X%**,
while OptiClust outperformed open reference OptiFit by **Y%**. **TODO vsearch
closed ref run time**.

For all datasets and clustering methods (_de novo_, open reference, and closed
reference), mothur's clustering algorithms produced higher quality OTUs than
vsearch.  In terms of runtime, OptiFit generally performed faster than vsearch
when reference clustering, while vsearch _de novo_ clustering outperformed
OptiClust.

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (**TODO: diff in MCC medians**). OTU quality was remarkably stable across
100 different random seeds. In terms of runtime, closed reference OptiFit
performed faster than OptiClust on whole datasets by **Z%**. In open reference
mode, OptiClust performed **X to Y%** faster than OptiFit only when the OptiFit
reference fraction was 30% or less. The split dataset strategy performed just as
well as the database strategy in open reference mode regardless of database
used, and outperformed the database strategy in closed reference mode by **W%**.

We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset. OTU
quality was similar with the simple and abundance-weighted sampling (median MCCs
**X** and **Y** respectively), but **Z%** worse with similarity-weighted
sampling. In closed reference mode, the fraction of query sequences that can be
fit to the reference OTUs decreases as the reference fraction increases from an
MCC of **A** with **J** reference sequences to an MCC of **B** with **K**
reference sequences.
